---
title: "Portfolio 6"
---

> The projects should be numbered consecutively (i.e., in the order in which you began them), and should include for each project a description of the goal, the product (computer program, hand graph, computer graph, etc.), the data, and some interpretation. Reports must be reproducible and of high quality in terms of writing, grammar, presentation, etc.

# Backgrounds 

This portfolio will follow the third goal we proposed in the prior portfolio. 

# Goals 

This portfolio will conduct secondary analyses on Target Fixation Rate 


# Secondary Analysis 3: Target Fixation Rate

The original analysis measured the inhibition side of attentional learning - how much the eye is suppressed from the singleton distractor. But attentional control involves two side - inhibiting the distractor and facilitating selection of the target. If oculomotor suppression of the singleton frees attention to be reallocated toward the target, then the rate at which first eye movements land on the target should increase across trials 1 to 3. 

In this way, this analysis asks: does suppression of the singleton distractor come with a corresponding enhancement of target selection? Finding both effects - singleton suppression and target enhancement would provide converging evidence that the 3-trial sequence produces a general retuning of the attentional priority map. 

One-sample t-tests against chance (1/6, given 6 equally-spaced items) test whether participants are above chance at fixating the target at each trial position. If target fixation is already above chance from trial 1, that suggests participants always have some target guidance; if it increases across trials, that suggests suppression actively redirects attention towards the target. 

**Predicted Patterns** 
- *Search task*: Target fixation rate increases from Trial 1 to Trial 3, as oculomotor resources freed from the singleton are redirected toward the target.
- *Central task*: Because only Trial 3 involves free eye movements, target fixation rate cannot change across trials in the same way; any apparent change should be treated cautiously.
- *Interaction*: A `newtrial × taskType` interaction would confirm that the increase in target selection is specific to the search task and driven by suppression learning.


## Data preparation 
We reuse the `firstfix2` structure from the original analysis pipeline, which already computes the percentage of first fixations to each item (Target, Singleton, Nonsingleton) by subject × task × trial. We simply extract the Target rows, which the original analysis never statistically tested.

```{r}
# Rebuild firstfix2 (identical to the original script)
accdata2 <- subset(accdata, !is.na(SRT))

firstfix2 <- as.data.frame(
  xtabs(~ curritem + subjnum + taskType + newtrial, data = accdata2)
)
firstfix2$total <- with(firstfix2,
  ave(Freq, interaction(subjnum, taskType, newtrial), FUN = sum))
firstfix2$perc  <- firstfix2$Freq / firstfix2$total
firstfix2$perc  <- with(firstfix2,
  ifelse(curritem == 'Nonsingleton', perc / 4, perc))
firstfix2$perc[is.nan(firstfix2$perc)] <- 0
firstfix2 <- firstfix2[, -c(5, 6)]

# Extract Target rows only — this is what the original script never tested
targfix <- firstfix2 %>% filter(curritem == "Target")

targfix %>%
  group_by(taskType, newtrial) %>%
  summarise(M  = round(mean(perc), 4),
            SD = round(sd(perc),   4),
            n  = n(), .groups = "drop") %>%
  kable(caption = "Table 3. Target first-fixation rate (proportion) by condition and trial position. Chance = 0.167 (1/6).")
```
In the search task, mean target fixation rate was 38.97% (SD = 14.79%) on trial 1, 39.72% (SD = 16.42) on trial 2, and 38.97% (SD = 16.72%) on trial 3, flat across the sequence and consistently above the 1/6 chance baseline. In the central task, only trial 3 yielded valid fixation data (eye movements were restricted on trials 1 and 2). Mean target fixation rate on central trial 3 was 37.76% (SD = 14.62%), close to the rates in search task at all trial positions. 

# 2(taskType) × 3(newtrial) within-subjects ANOVA

```{r}
MANOVA(
  data   = targfix,
  subID  = "subjnum",
  dv     = "perc",
  within = c('newtrial', 'taskType'),
  digits = 2
) %>%
  EMMEANS('newtrial') %>%
  EMMEANS('taskType') %>%
  EMMEANS('taskType', by = 'newtrial')
```


## One-way ANOVA within the search task

```{r}
targfix_search <- targfix %>% filter(taskType == "search")
MANOVA(
  data   = targfix_search,
  subID  = "subjnum",
  dv     = "perc",
  within = 'newtrial',
  digits = 2
) %>%
  EMMEANS('newtrial')
```

## One-sample t-tests vs. chance

These tests ask, at each trial position, whether participants' target fixation rate is reliably above the 1/6 chance baseline. This helps distinguish between two interpretations: (a) target guidance is always strong and does not meaningfully change, or (b) guidance is weak early (when the singleton competes) and strengthens across trials.

```{r}
chance <- 1 / 6
options(scipen = 999)

targfix_wide <- targfix_search %>%
  pivot_wider(names_from = newtrial, values_from = perc)

for (i in 1:3) {
  t_frame <- data.frame(perc = targfix_wide[[as.character(i)]])
  TTEST(t_frame, "perc", test.value = chance, paired = FALSE, digits = 3)
}
```

# Plot 

```{r targfix-plot, fig.width=9, fig.height=5, fig.cap="Figure S3. Target first-fixation rate (%) across trial positions by task type. Dashed gray line = chance (1/6 ≈ 16.7%). Error bars = 95% CI (Cousineau-Morey corrected)."}
targ_plot <- with(targfix,
                  cmErrBar(perc, subjnum, paste(taskType, newtrial, curritem), "ci")) %>%
  separate(group, into = c("condition", "trial", "item"), sep = " ")
targ_plot$trial <- factor(targ_plot$trial, levels = c("1", "2", "3"))

ggplot(targ_plot, aes(x = trial, y = mean, group = condition, color = condition)) +
  geom_errorbar(aes(ymin = minus, ymax = plus), width = 0.15, linewidth = 1.1) +
  geom_line(linewidth = 1.1, aes(linetype = condition)) +
  geom_point(size = 4, aes(shape = condition)) +
  geom_hline(yintercept = 1 / 6, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  annotate("text", x = 0.6, y = 1 / 6 + 0.012, label = "chance (1/6)",
           color = "gray50", size = 4.5, hjust = 0) +
  scale_color_manual(values = c("central" = "#4A90E2", "search" = "#333333"),
                     labels = c("Central Task", "Search Task")) +
  scale_shape_manual(values = c("central" = 17, "search" = 15),
                     labels = c("Central Task", "Search Task")) +
  scale_linetype_manual(values = c("central" = "dashed", "search" = "solid"),
                        labels = c("Central Task", "Search Task")) +
  scale_y_continuous(limits = c(0.1, 0.6), breaks = seq(0.1, 0.6, 0.1),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(x     = "Trials with Same Color",
       y     = "Target First-Fixation Rate",
       color = "Task Type", shape = "Task Type", linetype = "Task Type") +
  manuscript_theme
```


# Extension 1: Full First-Fixation Probability Decomposition

The main analysis only examined the target row of `firstfix2`. But understanding *where the freed probability mass goes* as singleton fixation drops is theoretically important. If suppression actively redirects attention toward the goal, target fixation should rise as singleton fixation falls. If suppression is purely inhibitory — reducing the singleton's pull without boosting target priority — target fixation should stay flat and the freed probability should flow instead to nonsingleton items.

This extension plots and tests all three item types (Singleton, Target, Nonsingleton per item) simultaneously across Trials 1–3 in the search task.
# Extension 2: Individual Differences — Does T1 Capture Predict Learning?

If being overtly captured by the singleton on Trial 1 is what drives suppression learning, then participants who showed stronger T1 capture should show stronger T2/T3 suppression. This extension tests that prediction directly at the individual level using Pearson correlations between each participant's Trial 1 singleton fixation rate and their subsequent learning magnitude (Trial 1 − Trial 3 reduction in singleton fixation).

