---
title: "Portfolio 11: HMM Stage 3 - Transfer Prediction (VST → RSVP)"
---

# Background

Portfolios 9 and 10 established that a 2-state joint HMM (item type + fixation duration) recovers two attentional modes - Efficient Search and Distractor Engagement - and that the VD-CD difference in state prevalence captures the learned distractor suppression signal. Portfolio 10 extracted five participant-level features from the VST-first group (n = 24) that quantify how quickly and how completely each individual learned to suppress the consistent distractor:

- Learning slope: rate of increase in P(Efficient Search) across trials.
- Transition trial: the trial at which the 5-trial rolling mean of P(Efficient Search) first exceeds 0.50.
- End-state P(Efficient Search): mean P(Efficient Search) in the final experimental block.
- Overall suppression score: mean VD-CD difference in P(Distractor Engagement) across blocks.
- Suppression growth: late - early change in the suppression score.

The central question for Stage 3 is whether these HMM-derived individual differences in VST suppression learning predict cross-task transfer to the RSVP task. The VST-first group completed the Visual Search Task before the RSVP task, so any relationship between VST learning features and RSVP outcomes would constitute evidence that the latent suppression process captured by the HMM generalizes beyond the training task.

The primary RSVP transfer outcome is Blink_CD - the attentional blink magnitude for the consistent distractor color, computed as the difference in target accuracy between lag 8 and lag 2, where lag is the distance in stream positions between a distractor (a hashtag) and the single target that follows it. If learned suppression transfers, participants who showed stronger or faster suppression learning in the VST should show a reduced attentional blink for the consistent distractor color in the RSVP task.

This approach follows the logic of individual-differences transfer designs used in the learned attention literature, where the strength of statistical learning during training predicts the magnitude of transfer effects in a novel task.


# Goals

1. Load the participant-level HMM features from Portfolio 10 and the RSVP transfer outcomes, and merge by participant.
2. Examine descriptive statistics and distributions of the RSVP outcome (Blink_CD) and all HMM predictors.
3. Run bivariate correlations between each HMM feature and Blink_CD to identify which aspects of suppression learning are associated with transfer.
4. Fit a multiple regression with the most promising predictors to test their joint and unique contributions.
5. Visualize the key predictor–outcome relationships.
6. Evaluate assumptions (normality, influential cases) and report results in APA style.


# Setup

```{r}
rm(list = ls())

library(tidyverse)
library(readxl)
library(knitr)
library(patchwork)
library(car)

select <- dplyr::select
filter <- dplyr::filter
```


```{r}
# Plot theme (matching manuscript style from p09/p10)
axis_text_size <- 14
axis_tick_text_size <- 12
graph_color <- "black"
axis_tick_length <- 0.25
axis_tick_units <- "cm"
axis_thickness <- 1

manuscript_theme <- theme_classic() +
  theme(
    axis.line = element_line(linewidth = axis_thickness, colour = graph_color),
    axis.title = element_text(size = axis_text_size),
    axis.text = element_text(colour = graph_color, size = axis_tick_text_size),
    axis.ticks = element_line(linewidth = axis_thickness, colour = graph_color),
    axis.ticks.length.y = unit(axis_tick_length, axis_tick_units),
    axis.ticks.length.x = unit(axis_tick_length, axis_tick_units),
    legend.position = "top",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    text = element_text(size = axis_tick_text_size)
  )

# Consistent color palette
col_efficient <- "#1F77B4"
col_engaged <- "#D62728"
```


# Data Loading and Merging

## Load HMM Features (from Portfolio 10)

```{r}
# Participant-level HMM features extracted in p10 (VST-first group, n = 24)
hmm_features <- read_csv("output/stage2_participant_features_vst_first.csv",
                         show_col_types = FALSE)
```


## Load RSVP Transfer Outcomes

```{r}
# RSVP accuracy data
rsvp_acc <- read_excel(
  "summary/RSVP_DataSummary_SL_EyeTracking_Final.xlsx",
  sheet = "ACC"
)

# Filter to VST-first group (Order == "VST_RSVP")
rsvp_vst <- rsvp_acc %>%
  filter(Order == "VST_RSVP") %>%
  mutate(subjnum = as.integer(subjnum))
```

Each RSVP stream contains a distractor (a hashtag) followed by a single target; lag indexes the number of stream positions between them. The dataset contains target accuracy for each condition (CD, ND, VD_Match, VD_Mismatch) at two lags (2 and 8), along with pre-computed attentional blink scores (Blink = target accuracy at lag 8 − target accuracy at lag 2). The primary outcome is Blink_CD: the attentional blink magnitude for the consistent distractor color.


## Merge Datasets

```{r}
# Merge by subject number. We use inner_join (not left_join) on purpose: any participant who has features but no RSVP record (or vice versa) is dropped from analysis, which keeps `transfer_data` to the participants who can contribute to every model below. The setdiff checks below report whether anyone was lost so a silent drop never goes unnoticed.
transfer_data <- hmm_features %>%
  inner_join(rsvp_vst, by = "subjnum")


# Diagnostic: who, if anyone, was lost in the merge
hmm_only <- setdiff(hmm_features$subjnum, rsvp_vst$subjnum)
rsvp_only <- setdiff(rsvp_vst$subjnum, hmm_features$subjnum)
if (length(hmm_only) > 0) cat("  Participants in HMM but not RSVP:", hmm_only, "\n")
if (length(rsvp_only) > 0) cat("  Participants in RSVP but not HMM:", rsvp_only, "\n")
if (length(hmm_only) == 0 & length(rsvp_only) == 0) cat("  All participants matched.\n")
```


# Descriptive Statistics

## RSVP Outcome: Blink_CD

```{r}
# Compute M/SD/Min/Max for each of the four blink measures, producing one wide row with columns like `Blink_CD_M`, `Blink_CD_SD`, ...
# The two pivots then reshape this into a tidy measure x {M,SD,Min,Max} table:
## pivot_longer's `names_pattern = "(.+)_(M|SD|Min|Max)"` splits each column name into a measure stem and a statistic suffix
## pivot_wider then puts each statistic in its own column
blink_summary <- transfer_data %>%
  summarise(
    across(c(Blink_CD, Blink_ND, Blink_VD_Match, Blink_VD_Mismatch),
           list(M = ~mean(., na.rm = TRUE),
                SD = ~sd(., na.rm = TRUE),
                Min = ~min(., na.rm = TRUE),
                Max = ~max(., na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(),
               names_to = c("measure", "stat"),
               names_pattern = "(.+)_(M|SD|Min|Max)") %>%
  pivot_wider(names_from = stat, values_from = value)

kable(blink_summary, digits = 3,
      caption = "Table 1. Descriptive statistics for RSVP attentional blink scores (VST-first group).")
```

```{r blink-distribution, fig.width=8, fig.height=5, fig.cap="Figure 1. Distribution of attentional blink magnitude for the consistent distractor color (Blink_CD). Each bar represents a VST-first participant. Positive values indicate a standard attentional blink (better target accuracy at lag 8 than lag 2); negative values indicate a reversed blink."}
ggplot(transfer_data, aes(x = Blink_CD)) +
  geom_histogram(binwidth = 0.05, fill = col_efficient, color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = mean(transfer_data$Blink_CD, na.rm = TRUE),
             linetype = "solid", color = col_engaged, linewidth = 1) +
  labs(x = "Blink_CD (Lag 8 - Lag 2 Accuracy)",
       y = "Count") +
  annotate("text", x = mean(transfer_data$Blink_CD, na.rm = TRUE), y = Inf,
           label = "Mean", vjust = 2, hjust = -0.2, color = col_engaged, size = 5) +
  manuscript_theme
```


## HMM Predictor Summary

```{r}
# Same wide-summarise -> pivot pattern as the blink summary above, applied to the five HMM predictors. We carry M and SD only (not Min/Max) because the predictors are bounded transformations of probabilities or trial indices and their range is less informative than the spread.
predictor_summary <- transfer_data %>%
  summarise(
    across(c(learning_slope, transition_trial, end_p_efficient,
             overall_suppress, suppress_growth),
           list(M = ~mean(., na.rm = TRUE),
                SD = ~sd(., na.rm = TRUE)),
           .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(),
               names_to = c("feature", "stat"),
               names_pattern = "(.+)_(M|SD)") %>%
  pivot_wider(names_from = stat, values_from = value)

kable(predictor_summary, digits = 4,
      caption = "Table 2. Descriptive statistics for HMM-derived predictors (VST-first group).")
```


# Bivariate Correlations

We first examine individual Pearson correlations between each HMM feature and Blink_CD. This step identifies which features show the strongest zero-order relationships with transfer before entering them into a multiple regression.

The expected sign of each correlation differs across features. Because a smaller blink (lower Blink_CD value) reflects better RSVP performance, stronger or faster suppression learning during the VST should produce a negative correlation with Blink_CD for `learning_slope`, `end_p_efficient`, `overall_suppress`, and `suppress_growth` - i.e., participants who learned more should have less blink. `transition_trial` runs in the opposite direction: faster learning yields a lower trial number, so the expected correlation with Blink_CD is positive (low transition trial paired with low blink). The same sign logic applies to the regression coefficients in the model below.

```{r bivariate-correlations}
# Predictor vector (used here and reused throughout the file) plus a named lookup for nicely formatted labels in tables and figures.
predictors <- c("learning_slope", "transition_trial", "end_p_efficient",
                "overall_suppress", "suppress_growth")

predictor_labels <- c(
  learning_slope = "Learning Slope",
  transition_trial = "Transition Trial",
  end_p_efficient = "End-State P(Efficient)",
  overall_suppress = "Overall Suppression",
  suppress_growth = "Suppression Growth"
)

# Run bivariate correlations with map_dfr (apply the function to each predictor name and row-bind the results into a tibble). 
cor_results <- map_dfr(predictors, function(pred) {
  d <- transfer_data %>%
    select(Blink_CD, all_of(pred)) %>%
    drop_na()

  ct <- cor.test(d[[pred]], d$Blink_CD)

  tibble(
    Predictor = predictor_labels[pred],
    variable = pred,
    r = ct$estimate,
    t = ct$statistic,
    df = ct$parameter,
    p = ct$p.value,
    CI_lower = ct$conf.int[1],
    CI_upper = ct$conf.int[2],
    n = nrow(d)
  )
})

kable(cor_results %>% select(Predictor, n, r, t, df, p, CI_lower, CI_upper),
      digits = c(0, 0, 3, 2, 0, 4, 3, 3),
      caption = "Table 3. Bivariate Pearson correlations between HMM features and Blink_CD.",
      col.names = c("Predictor", "n", "r", "t", "df", "p", "95% CI Lower", "95% CI Upper"))
```

```{r correlation-interpretation}
# Pre-compute summary counts for inline reporting in the prose below.
sig_predictors <- cor_results %>% filter(p < .05)
trend_predictors <- cor_results %>% filter(p >= .05 & p < .10)

n_sig <- nrow(sig_predictors)
n_trend <- nrow(trend_predictors)

# "Strongest" = predictor with the smallest p-value (which here also corresponds to the largest |r| up to df differences across predictors).
strongest <- cor_results %>% arrange(p) %>% slice(1)
```


## Scatterplots: Each HMM Feature × Blink_CD

```{r scatter-all, fig.width=12, fig.height=8, fig.cap="Figure 2. Bivariate relationships between each HMM feature and Blink_CD. Each point is one VST-first participant. The regression line (with 95% CI band) shows the direction and strength of the linear association. Stronger suppression learning during VST (e.g., steeper learning slope, higher end-state efficiency) should predict a smaller attentional blink for the consistent distractor color in the RSVP task."}
scatter_data <- transfer_data %>%
  select(subjnum, Blink_CD, all_of(predictors)) %>%
  pivot_longer(cols = all_of(predictors),
               names_to = "feature", values_to = "value") %>%
  mutate(feature_label = unname(predictor_labels[feature]))

ggplot(scatter_data, aes(x = value, y = Blink_CD)) +
  geom_point(size = 2.5, color = "gray35", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE,
              color = "black", fill = "gray80", linewidth = 1) +
  facet_wrap(~ feature_label, scales = "free_x", ncol = 3) +
  labs(x = "HMM Feature Value", y = "Blink_CD") +
  manuscript_theme +
  theme(strip.text = element_text(size = 11))
```


# Predictor Intercorrelations

Before fitting the multiple regression, we check intercorrelations among predictors to assess potential multicollinearity.

```{r predictor-intercorrelations}
# Compute the full predictor x predictor correlation matrix. 
# We drop_na() first to align all pairwise correlations on the same set of complete cases
# the `use = "pairwise.complete.obs"` argument is then a defensive default in case any NAs slip through. 
# High intercorrelations (|r| > .70 is a common rule of thumb) flag potential multicollinearity that would inflate VIFs and destabilise the unique-effect estimates in the multiple regression.
pred_matrix <- transfer_data %>%
  select(all_of(predictors)) %>%
  drop_na()

cor_matrix <- cor(pred_matrix, use = "pairwise.complete.obs")

kable(round(cor_matrix, 3),
      caption = "Table 4. Intercorrelation matrix among HMM predictors.")
```

```{r correlation-heatmap, fig.width=8, fig.height=7, fig.cap="Figure 3. Heatmap of intercorrelations among HMM predictors. Darker blue = stronger positive correlation; darker red = stronger negative correlation. High intercorrelations (|r| > .70) would signal multicollinearity concerns for the multiple regression."}
cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  rename(Var1 = Var1, Var2 = Var2, r = Freq) %>%
  mutate(
    Var1 = unname(predictor_labels[as.character(Var1)]),
    Var2 = unname(predictor_labels[as.character(Var2)])
  )

ggplot(cor_long, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5) +
  scale_fill_gradient2(low = col_engaged, mid = "white", high = col_efficient,
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10))
```


# Multiple Regression: HMM Features Predicting Blink_CD

We fit a multiple regression entering all five HMM features as simultaneous predictors of Blink_CD. This tests whether the features carry unique predictive information about transfer beyond their shared variance. With *n* = 24 and 5 predictors, the model has only 18 residual degrees of freedom; estimates of unique predictor contributions are therefore underpowered, and small-to-moderate true effects may not reach significance even if they exist. The standardized coefficients reported below should be read alongside the bivariate correlations rather than as a definitive test of any single feature's predictive value.

```{r multiple-regression}
# Fit the full model (listwise deletion for transition_trial NAs)
reg_data <- transfer_data %>%
  select(subjnum, Blink_CD, all_of(predictors)) %>%
  drop_na()

model_full <- lm(Blink_CD ~ learning_slope + transition_trial +
                   end_p_efficient + overall_suppress + suppress_growth,
                 data = reg_data)

summary(model_full)
```


## Regression Diagnostics

```{r regression-diagnostics}
# Variance Inflation Factors
vif_values <- vif(model_full)
print(round(vif_values, 2))

vif_concern <- any(vif_values > 5)
if (vif_concern) {
  cat("\nWARNING: VIF > 5 detected - multicollinearity may inflate standard errors.\n")
} else {
  cat("\nAll VIFs < 5 - no severe multicollinearity concerns.\n")
}
```

```{r diagnostic-plots, fig.width=10, fig.height=8, fig.cap="Figure 4. Regression diagnostic plots. Top-left: residuals vs. fitted (check for non-linearity and heteroscedasticity). Top-right: Q-Q plot (check for normality of residuals). Bottom-left: scale-location (check for homoscedasticity). Bottom-right: residuals vs. leverage (identify influential cases with high Cook's distance)."}
# `plot.lm()` returns four diagnostic panels in sequence; par(mfrow) lays
# them out in a 2x2 grid so they render as one figure. Reset to single-panel
# afterwards so subsequent base-R plots are not affected.
par(mfrow = c(2, 2))
plot(model_full)
par(mfrow = c(1, 1))
```

```{r influential-cases}
# Cook's D > 4/n is a widely used cutoff (Bollen & Jackman, 1990 - but cited
# only as a heuristic here, not a strict rule). It flags observations whose
# leverage * residual product is large enough to materially shift the fitted
# coefficients. With n = 24 the threshold is generous (~0.17), so a flagged
# case warrants inspection rather than automatic removal.
cooks_d <- cooks.distance(model_full)
threshold <- 4 / nrow(reg_data)
influential <- which(cooks_d > threshold)

if (length(influential) > 0) {
  cat(sprintf("Influential cases (Cook's D > %.3f):\n", threshold))
  cat("  Row indices:", paste(influential, collapse = ", "), "\n")
  cat("  Subject numbers:", paste(reg_data$subjnum[influential], collapse = ", "), "\n")
  cat("  Cook's D values:", paste(round(cooks_d[influential], 3), collapse = ", "), "\n")
} else {
  cat("No influential cases detected.\n")
}
```


## Standardized Coefficients

```{r standardized-regression}
# Standardize all variables (X and Y) for comparable effect sizes; the
# resulting lm coefficients are standardized betas. We wrap scale() in
# as.numeric() because scale() returns a 1-column matrix with attributes,
# which dplyr would store as a matrix column - fine for lm() in most cases
# but fragile for downstream broom::tidy() across versions.
reg_data_z <- reg_data %>%
  mutate(across(c(Blink_CD, all_of(predictors)), ~ as.numeric(scale(.))))

model_std <- lm(Blink_CD ~ learning_slope + transition_trial +
                  end_p_efficient + overall_suppress + suppress_growth,
                data = reg_data_z)

# Extract standardized coefficients
std_coefs <- broom::tidy(model_std) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = unname(predictor_labels[term]))

kable(std_coefs, digits = c(0, 3, 3, 2, 4),
      caption = "Table 5. Standardized regression coefficients predicting Blink_CD from HMM features.",
      col.names = c("Predictor", "β", "SE", "t", "p"))
```

```{r coefficient-plot, fig.width=8, fig.height=5, fig.cap="Figure 5. Standardized regression coefficients (β) with 95% confidence intervals. Bars crossing zero indicate non-significant predictors at α = .05."}
# Reorder the predictor factor by |β| so the largest-magnitude coefficient
# sits at the top of the y-axis and the weakest at the bottom - this puts
# the most informative predictors visually first regardless of sign.
std_coefs_ci <- broom::tidy(model_std, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = unname(predictor_labels[term]),
         term = fct_reorder(term, abs(estimate)))

ggplot(std_coefs_ci, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8) +
  geom_point(size = 3, color = col_efficient) +
  labs(x = "Standardized Coefficient (β)",
       y = NULL) +
  manuscript_theme
```


# Exploratory: Alternative RSVP Outcomes

As an exploratory extension, we examine whether HMM features also predict the attentional blink for other distractor conditions (Blink_ND, Blink_VD_Match, Blink_VD_Mismatch). If the HMM-captured suppression is specific to the consistent distractor color, we would expect stronger prediction of Blink_CD than of other blink measures.

```{r exploratory-outcomes}
# Same map_dfr pattern as the primary bivariate-correlations chunk, but inverted: instead of iterating over predictors with one outcome (Blink_CD), we iterate over outcomes with one predictor (the strongest from the primary analysis). 
# The expected pattern under CD-specific transfer is that the correlation magnitude is largest for Blink_CD and smaller for the other blink measures, where there was no opportunity for VST learning to help.
outcomes <- c("Blink_CD", "Blink_ND", "Blink_VD_Match", "Blink_VD_Mismatch")
outcome_labels <- c(
  Blink_CD = "Consistent Distractor",
  Blink_ND = "No Distractor",
  Blink_VD_Match = "Variable (Match)",
  Blink_VD_Mismatch = "Variable (Mismatch)"
)

best_pred <- strongest$variable

exploratory_cors <- map_dfr(outcomes, function(outcome) {
  d <- transfer_data %>%
    select(all_of(outcome), all_of(best_pred)) %>%
    drop_na()

  ct <- cor.test(d[[best_pred]], d[[outcome]])

  tibble(
    Outcome = outcome_labels[outcome],
    Predictor = predictor_labels[best_pred],
    r = ct$estimate,
    p = ct$p.value,
    n = nrow(d)
  )
})

kable(exploratory_cors, digits = c(0, 0, 3, 4, 0),
      caption = sprintf("Table 6. Correlation between %s and each RSVP blink measure.",
                        predictor_labels[best_pred]))
```


# Specificity Analysis: Blink Cost (CD - VD_Mismatch)

The primary analysis above tested whether HMM features predict the absolute attentional blink for the consistent distractor color (Blink_CD). However, Blink_CD confounds the trained-distractor effect with general attentional ability. To quantify the transfer signal relative to a baseline distractor that participants had no opportunity to learn about, we compute a blink cost score: Blink_Cost = Blink_CD - Blink_VD_Mismatch. A negative cost indicates that the consistent distractor color produces *less* attentional blink than a variable-location distractor with a non-matching colour - consistent with learned suppression transferring to the RSVP task and reducing the attentional cost of the trained colour.

If HMM features predict Blink_CD but not Blink_Cost, the apparent association reflects general attentional efficiency that lifts target accuracy across all conditions. If they predict Blink_Cost, that provides stronger evidence for transfer specific to the trained colour relative to an untrained-colour baseline.

```{r blink-cost-compute}
# Compute blink cost: CD - VD_Mismatch (untrained-colour baseline).
# Negative = less blink for the trained CD colour than for an unmatched variable distractor - the expected direction if learned suppression transfers.
transfer_data <- transfer_data %>%
  mutate(Blink_Cost = Blink_CD - Blink_VD_Mismatch)

# One-sample t-test: is blink cost different from zero?
t.test(transfer_data$Blink_Cost)
```

```{r blink-cost-distribution, fig.width=6, fig.height=4, fig.cap="Figure 6. Distribution of blink cost scores (Blink_CD - Blink_VD_Mismatch). Negative values indicate that the consistent distractor colour produces less attentional blink than an unmatched-colour variable distractor - consistent with suppression transfer to the trained colour."}
ggplot(transfer_data, aes(x = Blink_Cost)) +
  geom_histogram(fill = col_engaged, alpha = 0.7, bins = 12,
                 color = "white", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  geom_vline(xintercept = mean(transfer_data$Blink_Cost, na.rm = TRUE),
             color = col_efficient, linewidth = 1.2) +
  labs(x = "Blink Cost (CD - VD_Mismatch)",
       y = "Count") +
  manuscript_theme
```

## Bivariate Correlations: HMM Features × Blink Cost

```{r blink-cost-correlations}
# Run bivariate correlations with Blink_Cost
cost_cors <- map_dfr(predictors, function(pred) {
  d <- transfer_data %>%
    select(all_of(pred), Blink_Cost) %>%
    drop_na()

  ct <- cor.test(d[[pred]], d$Blink_Cost)

  tibble(
    Predictor = predictor_labels[pred],
    n = nrow(d),
    r = ct$estimate,
    t = ct$statistic,
    df = ct$parameter,
    p = ct$p.value,
    CI_lower = ct$conf.int[1],
    CI_upper = ct$conf.int[2]
  )
})

kable(cost_cors, digits = c(0, 0, 3, 2, 0, 4, 3, 3),
      caption = "Table 7. Bivariate correlations between HMM features and Blink Cost (CD - VD_Mismatch).")
```

```{r blink-cost-scatterplots, fig.width=10, fig.height=8, fig.cap="Figure 7. Scatterplots of each HMM feature against Blink Cost. Blue lines show least-squares fit with 95% CI."}
plot_data_cost <- transfer_data %>%
  select(all_of(predictors), Blink_Cost) %>%
  pivot_longer(cols = all_of(predictors), names_to = "feature", values_to = "value") %>%
  mutate(feature = unname(predictor_labels[feature]),
         feature = factor(feature, levels = unname(predictor_labels)))

ggplot(plot_data_cost, aes(x = value, y = Blink_Cost)) +
  geom_point(size = 2.5, color = "gray35", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = col_efficient, fill = col_efficient, alpha = 0.2) +
  facet_wrap(~ feature, scales = "free_x", ncol = 3) +
  labs(x = "Feature Value", y = "Blink Cost (CD - VD_Mismatch)") +
  manuscript_theme +
  theme(strip.text = element_text(size = 11))
```

## Multiple Regression: HMM Features Predicting Blink Cost

```{r blink-cost-regression}
# Fit the full model
cost_reg_data <- transfer_data %>%
  select(subjnum, Blink_Cost, all_of(predictors)) %>%
  drop_na()

model_cost <- lm(Blink_Cost ~ learning_slope + transition_trial +
                   end_p_efficient + overall_suppress + suppress_growth,
                 data = cost_reg_data)

summary(model_cost)

# Extract key statistics
cost_r2 <- summary(model_cost)$r.squared
cost_adj_r2 <- summary(model_cost)$adj.r.squared
cost_f <- summary(model_cost)$fstatistic
cost_f_p <- pf(cost_f[1], cost_f[2], cost_f[3], lower.tail = FALSE)

cat(sprintf("\nBlink Cost model: R² = %.3f, adj R² = %.3f, F(%d, %d) = %.2f, p = %.4f\n",
            cost_r2, cost_adj_r2, cost_f[2], cost_f[3], cost_f[1], cost_f_p))

# Pre-compute Blink_CD model stats for inline text below
r2 <- summary(model_full)$r.squared
adj_r2 <- summary(model_full)$adj.r.squared
f_stat <- summary(model_full)$fstatistic
f_p <- pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE)
```

```{r blink-cost-comparison, fig.width=9, fig.height=6, fig.cap="Figure 8. Comparison of bivariate correlations: HMM features predicting Blink_CD (general blink) vs. Blink Cost (CD - VD_Mismatch, transfer-specific). Larger absolute correlations with Blink_CD than Blink Cost would suggest general attentional efficiency drives the association rather than transfer to the trained colour."}
# Side-by-side comparison of effect sizes for each predictor against the two outcomes: raw Blink_CD vs. Blink Cost (CD - VD_Mismatch). Same point styling encodes significance (filled triangle = p < .05, filled circle = p < .10, open circle = n.s.); position_dodge separates the two outcomes per predictor.
comparison <- bind_rows(
  cor_results %>%
    select(Predictor, r, p) %>%
    mutate(Outcome = "Blink_CD"),
  cost_cors %>%
    select(Predictor, r, p) %>%
    mutate(Outcome = "Blink Cost\n(CD - VD_Mismatch)")
) %>%
  mutate(sig = case_when(p < .05 ~ "p < .05", p < .10 ~ "p < .10", TRUE ~ "n.s."),
         Predictor = factor(Predictor, levels = rev(unname(predictor_labels))))

ggplot(comparison, aes(x = r, y = Predictor, color = Outcome, shape = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Blink_CD" = col_efficient,
                                "Blink Cost\n(CD - VD_Mismatch)" = col_engaged)) +
  scale_shape_manual(values = c("p < .05" = 17, "p < .10" = 16, "n.s." = 1)) +
  labs(x = "Pearson r", y = NULL, color = "Outcome", shape = "Significance") +
  manuscript_theme +
  theme(legend.position = "bottom")
```

`r if(cost_f_p >= .05 & f_p >= .05) sprintf("Neither the Blink_CD model (R² = %.3f, p = %.4f) nor the Blink Cost model (R² = %.3f, p = %.4f) reached significance. The pattern of bivariate correlations can clarify whether the trend-level effects observed for Blink_CD reflect general efficiency or transfer to the trained colour.", r2, f_p, cost_r2, cost_f_p) else if(cost_f_p < .05) sprintf("The Blink Cost model was significant (R² = %.3f, p = %.4f), providing evidence that HMM features predict transfer to the trained distractor colour beyond general attentional efficiency.", cost_r2, cost_f_p) else sprintf("The Blink_CD model (R² = %.3f, p = %.4f) was significant but the Blink Cost model was not (R² = %.3f, p = %.4f), suggesting the predictive signal reflects general attentional efficiency rather than transfer to the trained colour.", r2, f_p, cost_r2, cost_f_p)`


# Summary

The multiple regression testing whether HMM-derived features from the VST predict attentional blink magnitude for the consistent distractor color (Blink_CD) in the RSVP task yielded R² = `r sprintf("%.3f", r2)` (adjusted R² = `r sprintf("%.3f", adj_r2)`), F(`r f_stat[2]`, `r f_stat[3]`) = `r sprintf("%.2f", f_stat[1])`, p = `r sprintf("%.4f", f_p)`.

`r if(f_p < .05) "The overall model was significant, indicating that HMM-derived suppression learning features carry meaningful predictive information about cross-task transfer. This supports the view that learned distractor suppression - as captured by the latent state dynamics of the HMM - generalizes beyond the training task to influence attentional processing in a qualitatively different paradigm." else "The overall model did not reach significance, suggesting that the HMM-derived suppression learning features did not reliably predict attentional blink magnitude for the consistent distractor color in this sample. This null result may reflect several factors: the modest sample size (n = 24) limits power to detect moderate effects; the RSVP blink measure may capture processes partially distinct from the spatial suppression indexed by the HMM; or transfer of learned suppression may operate through mechanisms not fully captured by the five features extracted here. Nonetheless, the direction and magnitude of individual predictor effects remain informative for guiding future work with larger samples."`

At the bivariate level, `r if(n_sig > 0) sprintf("%d predictor(s) showed significant correlations with Blink_CD (p < .05): %s.", n_sig, paste(sig_predictors$Predictor, collapse = ", ")) else "no individual predictor reached significance at p < .05."` `r if(n_trend > 0) sprintf("%d predictor(s) showed trends (p < .10): %s.", n_trend, paste(trend_predictors$Predictor, collapse = ", ")) else ""` The strongest zero-order relationship was with `r strongest$Predictor` (r = `r sprintf("%.3f", strongest$r)`, p = `r sprintf("%.4f", strongest$p)`).

## Specificity: Blink Cost (CD - VD_Mismatch)

To test whether the HMM features predict transfer to the trained distractor colour rather than general attentional efficiency, we computed a blink cost score (Blink_CD - Blink_VD_Mismatch) that benchmarks the trained-colour blink against a variable distractor with a non-matching colour - a baseline that participants had no opportunity to learn about. `r if(cost_f_p < .05) sprintf("The Blink Cost regression was significant (R² = %.3f, p = %.4f), supporting transfer specific to the trained distractor colour beyond what general attentional efficiency would explain.", cost_r2, cost_f_p) else sprintf("The Blink Cost regression was not significant (R² = %.3f, p = %.4f).", cost_r2, cost_f_p)` `r if(cost_f_p >= .05 & f_p >= .05) "Neither the raw Blink_CD model nor the Blink Cost model reached significance, consistent with the overall null transfer finding." else if(cost_f_p >= .05 & f_p < .05) "The fact that Blink_CD was predicted but Blink Cost was not suggests the HMM features capture general attentional efficiency rather than transfer specific to the trained colour." else ""`

The comparison of bivariate correlation patterns across the two outcomes (Figure 8) provides the clearest diagnostic: if correlations weaken when moving from Blink_CD to Blink Cost, the original associations were partially driven by general capacity differences rather than suppression-specific transfer. This distinction is important for interpreting the theoretical significance of any HMM-RSVP relationship.


